---
#title: "README"
#author: "Marco Rozgic"
#date: "3/21/2020"
output: html_document
pagetitle: "COVID-19 outbreak analysis in Germany with R"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pathData = "/Users/marcorozgic/Documents/FunRStuff/CovidAnalysis"
library(tidyverse)
#md_document
# package to handle date data types, which are allways hell
library(lubridate)

# this package has some nice themes for graphics which I find appealing
library(see)

# allows for great annotaions using our data making the code transferable
library(glue)

# nice package to arrange multiple plots
library(patchwork)

# helps us to make interactive plots
library(plotly)

# make even nicer plots
library(ggforce)

# a package containing an ODE solver for the SIR equation
library(deSolve)

# epidemiology package
library(EpiEstim)

#######
source(paste0(pathData,"/humanNumbers.R"))
#######
```

# COVID-19 outbreak analysis in Germany with R

The conducted analysis follows the posts by Tim Churches, starting with his post in February 2020:
<https://timchurches.github.io/blog/posts/2020-02-18-analysing-covid-19-2019-ncov-outbreak-data-with-r-part-1/#estimating-changes-in-the-effective-reproduction-number>

The code is slightly changed, some graphs are tweaked. All in all this is supposed to help scientists as well as non-scientists to gain insights and conduct their own analysis of the situation. The main code is found in **covidAnalysis.R**. I will try to show some results here on the main page. Please, feel free to contribute.


## Disclaimer
I am **not** a medical doctor, I am only a data-dude who wants to help citizen data scientist to stay informed and check the numbers we are confronted with every day. 
If you,  on the other hand, are someone who understands more about epidemiology, feel  free to use all you find here. Remember, most of it is presented in a much better way by Tim Churches.


## Data Acquisition
Data are pulled from JHU Git Hub archive <https://raw.githubusercontent.com/CSSEGISandData/>.


```{r read_data, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
jhu_url = paste("https://raw.githubusercontent.com/CSSEGISandData/",
                 "COVID-19/master/csse_covid_19_data/", "csse_covid_19_time_series/",
                 "time_series_19-covid-Confirmed.csv", sep = "")

dat =  read_csv(jhu_url) %>% rename(province = "Province/State",
                                    country = "Country/Region")

datWorld =  dat %>% pivot_longer(-c(province,country, Lat, Long),
                          names_to = "Date", 
                          values_to = "cumulative_cases")
europe = read_csv2(file =paste0(pathData, "/countriesOfEurope.csv"))

# consider "Mainland" only for the time beeing
datEurope = filter(datWorld, country %in% europe$Countries,
                   (province %in% europe$Countries | is.na(province)))

tPop = readRDS(file = paste0(pathData, "/tPop.RDS"))
cCodes = readRDS(file = paste0(pathData, "/countryCodes.RDS"))

datEurope = datEurope %>% 
  select (-province) %>% 
  left_join(., cCodes %>%  select(name, charcode), 
            by = c("country"="name")) %>% 
  left_join(., tPop %>% 
              filter(Year =="2020") %>%
              select(-Year),
            by ="charcode") %>% 
  mutate(Date = mdy(Date), value = value * 1000)


datEurope = datEurope %>% filter(cumulative_cases > 100) %>% 
  group_by(country) %>%
  group_modify(~{
    .x$daySince100 = 1:nrow(.x)
    return(.x)
  })%>% ungroup()


datLong = dat %>%
  filter(country == "Germany") %>%  
  pivot_longer(-c(province,country, Lat, Long),
               names_to = "Date",
               values_to = "cumulative_cases")
datLong = datLong %>%
  select(-c(province, Lat, Long)) %>%
  mutate(Date = mdy(Date))
datLong = datLong %>%  
  filter(cumulative_cases != 0) %>%
  mutate(incident_cases = c(0, diff(cumulative_cases)),
         myDay = 1:nrow(.),
         myWeek = week(Date)-3 )

p1 = datLong %>% ggplot(aes(x=Date, y = cumulative_cases)) +
  geom_line() +
  geom_point(shape = "x") +
  labs(title = glue("Number of COVID-19 infections in {datLong$country[1]}"),
       subtitle = glue("data ranging from {min(datLong$Date)} to {max(datLong$Date)}")
       ) +
  theme_lucid()+
  scale_fill_material_c()
```

For Germany the curve for example looks like this:

```{r echo=FALSE}
p1
```

Some people are interested in the growth of cases after for examople the 100th case. In most cases I see those figures with absolute numbers. I think this is kind of missleading. Using the the **wppExplorer** package I was able to obtain the total number of people in each european country.

### Development of cases in european countries after reaching 100 infections -Only mainland provinces are considered (e.g. only France not St.Martin)

```{r echo=FALSE, warning=FALSE}
(datEurope %>% 
  ggplot(aes(x= daySince100, y = (cumulative_cases/value)*100, colour = country, group = country) )+
           geom_line() +
           labs(x = "Number of days after reaching 100 cases",
                y =  "Percentage of population per country"
                ) +
           theme_lucid() +
  scale_colour_flat_d() )%>% ggplotly()
```


## Implemented Features
As to this day `r lubridate::today()`only simple models are considered.


### Linear modelling
A simple liner model is fit can easily be fitted to the log-transformed data:


```{r linModel, echo=TRUE}
myLinearModel = lm(log(cumulative_cases) ~ myDay,
                   datLong %>%
                   filter(Date >= as.Date("2020-02-24"))
                   )
summary(myLinearModel)
```
```{r linModelPlot, echo=FALSE}
linModelDf = broom::tidy(myLinearModel)
datLong %>% filter(Date >= as.Date("2020-02-24") ) %>%
  ggplot(aes(x=myDay, y= log(cumulative_cases))) +
  geom_smooth(method = lm) +
  geom_point() +
  labs(title = glue("fitted linear model with  intercept {round(linModelDf$estimate[1],2)} and slope {round(linModelDf$estimate[2],2)}"))+
  theme_lucid()

```

With a linear model crude predictions can be made:
```{r predictions, message=FALSE, warning=FALSE, include=FALSE}
startDay =  datLong %>%
  filter(Date >= as.Date("2020-02-24")) %>% pull(myDay) %>% min()
endDay = datLong %>%
  filter(Date >= as.Date("2020-02-24")) %>% pull(myDay) %>% max()

startDate = datLong %>%
  filter(Date >= as.Date("2020-02-24")) %>% pull(Date) %>% min()

endDate = datLong %>%
  filter(Date >= as.Date("2020-02-24")) %>% pull(Date) %>% max()

predDF = broom::tidy(predict(myLinearModel,
                             newdata = data.frame(myDay = startDay:(endDay+7)),
                             interval = "prediction"))

predDF$Date = seq(startDate,
                  max(datLong$Date)+days(7),
                  by = "day")


p4 = p1 + geom_ribbon(data = predDF,
                 inherit.aes = FALSE,
                 aes(ymin=exp(lwr), ymax = exp(upr), x = Date),
                 fill = "grey2", alpha =0.25) +
  geom_line(data = predDF,  inherit.aes = FALSE,
            aes(x= Date, y= exp(fit), colour = "fit")) +
  annotate("text",
           y = c(max(exp(predDF$fit)), max(exp(predDF$lwr)),max(exp(predDF$upr))),
           x = max(predDF$Date)+days(2),
           label = c(glue({round(max(exp(predDF$fit)))}),
                     glue({round(max(exp(predDF$lwr)))}),
                     glue({round(max(exp(predDF$upr)))})
                     )
           )+
  labs(title = glue("Number of COVID-19 infections in {datLong$country[1]} with a 7 day fit") )+
  theme(legend.title = element_blank())
```
```{r picPred, echo=FALSE}
p4 + facet_zoom(xlim= c(endDate, max(predDF$Date)))
```



### SIR modelling
A more sophisticated way to model the outbreak can be performed by applying the SIR-Model
(**S**usceptible **I**nfectious **R**ecovered). The model is based on an ODE system. See the code and Tim Churches' posts for more details. However, the following can be obtained:
```{r include=FALSE, paged.print=TRUE}

numPeople = 1000*83149.3
sirStartDate = min(datLong$Date)


init = c(S= numPeople - datLong$cumulative_cases[1],
         I = datLong$cumulative_cases[1],
         R = 0)

# first we need a function which translates the ODE's stated in the SIR model

SIR = function(time, state, parameters) {
  par = as.list(c(state, parameters))
  with(par, {
    dS = -beta * I * S/numPeople
    dI = beta * I * S/numPeople - gamma * I
    dR = gamma * I
    return(list(c(dS, dI, dR)))
  })
}

RSS = function(parameters) {
  names(parameters) = c("beta", "gamma")
  out = ode(y = init, times = datLong$myDay, func = SIR, parms = parameters)
  fit = out[, 3]
  return(sum((datLong$cumulative_cases - fit)^2))
}


Opt = optim(c(0.5, 0.5),
             RSS,
             method = "L-BFGS-B",
             lower = c(0,0),
             upper = c(1, 1)
             )
t = 1:100

optParams = set_names(Opt$par, "beta", "gamma")

fittedCumulativeIncidenceDf = data.frame(ode(y = init,
                                             times = t,
                                              func = SIR,
                                             parms = optParams)) %>%
  mutate(Date = min(datLong$Date) + days(time-1))


dat = left_join(fittedCumulativeIncidenceDf, datLong ,by=c("Date"))


```
```{r SIRPic, echo=FALSE, warning=FALSE}

dat %>% select(Date, cumulative_cases, I) %>%
  rename(cCases =cumulative_cases, fittedCCases = I) %>%
  pivot_longer(-Date) %>%
  ggplot(aes(x=Date, y=value, colour = name)) +
  geom_line() +
  geom_point(shape = "x") +
  labs(title = glue("SIR-Model based projection of COVID-19 infections in {datLong$country[1]}"),
       subtitle = glue("data ranging from {min(dat$Date)} to {max(dat$Date)}"),
       caption = glue("fitted data from SIR model,\n
                      computed R0 ={round(optParams[1]/optParams[2],2)},\n
                      fitted point of reduce: At {dat$Date[which(dat$I==max(dat$I))]} with {humanNumbers(round(dat$I[which(dat$I==max(dat$I))]))} infections"),
       y = "Cumulative Cases") +

  scale_y_continuous(labels = humanNumbers)+
  theme_lucid()+
  scale_color_material_d() +
  facet_zoom( y = name =="cCases",
              xlim = c(min(datLong$Date), max(datLong$Date)),
              horizontal = FALSE)
```


### 






<!-- ### Estimating Changing Reproduction Numbers -->
<!-- With  the EpiEstim package changing reproduction rates can be computed. So far only constant changerates have been considered. -->


